---
title: 'AFL: wins and losses after the bye'
author: "Neil Saunders"
date: "`r Sys.time()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(readxl)
library(broom)
library(lubridate)
library(fitzRoy)
library(pander)

theme_set(theme_bw())

```

# Introduction
Geelong have not won a game after the bye since 2011. Is this unusual? Can we shed any light as to why?

# Data
Our data comes from two sources:

- `fitzRoy::get_match_results()`
- AFL historical odds in Excel format [available from here](http://www.aussportsbetting.com/data/historical-afl-results-and-odds-data/)

```{r get-data, cache=TRUE}
results <- get_match_results()

afl_odds <- read_excel("../../../../data/afl.xlsx", 
    col_types = c("date", "text", "text", 
        "text", "text", "numeric", "numeric", 
        "text", "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "text"), skip = 1) %>% 
  select(Date, Home.Team = `Home Team`, Away.Team = `Away Team`, Venue, Home.Odds = `Home Odds`, Away.Odds = `Away Odds`)
```

# Processing
First question: when asking whether wins/losses after the bye are unusual, we need to ask "compared with what?"

For this study we'll compare the round before the bye with the week after.

```{r results-to-bye}
results_bye <- results %>% 
  select(Season, Round, Date, Venue, Home.Team, Away.Team, Margin) %>% 
  gather(Status, Team, -Season, -Round, -Margin, -Date, -Venue) %>% 
  filter(Season > 2010, grepl("^R", Round)) %>% 
  separate(Round, into = c("prefix", "suffix"), sep = 1) %>% 
  mutate(suffix = as.numeric(suffix)) %>% 
  arrange(Season, Team, suffix) %>%
  group_by(Season, Team) %>% 
  mutate(bye = case_when(
    suffix - lead(suffix) == -2 ~ "before",
    suffix - lag(suffix) == 2 ~ "after",
    TRUE ~ as.character(suffix)
  ),
  Margin = ifelse(Status == "Away.Team", -Margin, Margin)) %>% 
  ungroup() %>% 
  filter(bye %in% c("before", "after")) %>% 
  mutate(Result = case_when(
    Margin > 0 ~ "W",
    Margin < 0 ~ "L",
    TRUE ~ "D"
  )) %>% 
  unite(Round, prefix, suffix, sep = "")
```

A very long dplyr/tidyr chain gives us the relevant games. `r results_bye %>% distinct(Date, Venue) %>% nrow()` games have been played either side of a bye since 2011. Of these, `r results_bye %>% group_by(Date, Venue) %>% filter(n() == 1) %>% nrow()` games involved one team in the bye, and `r results_bye %>% group_by(Date, Venue) %>% filter(n() == 2) %>% nrow()/2` involved both teams in the bye.

# Analysis

## Visualisation
Let's confirm that Geelong have indeed not won a game after a bye since 2011:

```{r plot-win-loss-01}
results_bye %>% 
  filter(Team == "Geelong") %>% 
  pander(split.table = Inf,
         justify = c("right", "left", "left", "left", "right", "left", "left", "left", "left"))
```

Various ways to visualise this:

```{r plot-win-loss-02}
results_bye %>% 
  count(Team, Status, bye, Result) %>% 
  mutate(bye = factor(bye, levels = c("before", "after"))) %>% 
  ggplot(aes(bye, n)) + 
  geom_col(aes(fill = Result)) + 
  facet_wrap(~Team, ncol = 6) +
  scale_fill_manual(values = c("yellow2", "steelblue3"))
```

## Fisher exact test
We'll try a Fisher exact test to compare the win/loss count before the bye with the count after the bye.

The result is significant for Geelong alone at _p_ < 0.05, albeit with a very wide confidence interval.

```{r run-fisher-test}
results_bye %>% 
  count(Team, bye, Result) %>% 
  nest(-Team) %>% 
  mutate(data = map(data, . %>% spread(Result, n) %>% select(2:3)), 
         fisher = map(data, fisher.test), 
         summary = map(fisher, tidy)) %>% 
  select(Team, summary) %>% 
  unnest() %>% 
  select(-method, -alternative) %>% 
  arrange(p.value) %>% 
  pander(split.table = Inf)
```

## Visualisation by home/away
Splitting the results further by home/away is interesting: Geelong has lost all away games that occurred after a bye.

The only comparable team in that respect is Port Adelaide. However, other teams have struggled to win away _before_ a bye (Brisbane), or _at home before_ a bye (Carlton, North Melbourne). Of course, some teams struggle to win regardless of venue, opponent or proximity to a bye (Gold Coast).

```{r plot-win-loss-03}
results_bye %>% 
  count(Team, Status, bye, Result) %>% 
  mutate(i = interaction(Status, bye, lex.order = TRUE)) %>% 
  ggplot(aes(i, n)) + 
  geom_col(aes(fill = Result)) +
  facet_wrap(~Team, ncol = 6) + 
  coord_flip() + 
  scale_y_continuous(breaks = seq(1, 8, 2)) +
  scale_fill_manual(values = c("yellow2", "steelblue3"))
```

Note that several of the lost post-bye away games were at venues traditionally viewed as being difficult to win at for an away team (Adelaide, Perth):

```{r away-losses-01}
results_bye %>% 
  filter(Team == "Geelong", bye == "after", Status == "Away.Team") %>% 
  pander(split.table = Inf,
         justify = c("right", "left", "left", "left", "right", "left", "left", "left", "left"))
```

Also, three of the losses were against a side also coming off the bye, but playing at home:

```{r away-losses-02}
results_bye %>% 
  group_by(Date, Venue) %>% 
  filter(n() == 2) %>% 
  ungroup() %>% 
  filter(Team == "Geelong", bye == "after") %>% 
  pander(split.table = Inf,
         justify = c("right", "left", "left", "left", "right", "left", "left", "left", "left"))
```

For comparison, the away games before byes:

```{r away-losses-03}
results_bye %>% 
  filter(Team == "Geelong", bye == "before", Status == "Away.Team") %>% 
  pander(split.table = Inf,
         justify = c("right", "left", "left", "left", "right", "left", "left", "left", "left"))
```

## Odds
We might ask of the games lost after a bye - how many were Geelong expected to win?

For this we join the dataset with historical betting odds. Their accuracy is taken at face value.

Note that Geelong have never won after a bye when expected to lose. Of their 8 losses after the bye, 2 were expected and 6 were not.
```{r compare-expected-odds, message=FALSE}
afl_odds %>% 
  gather(Status, Team, 2:3) %>% 
  filter(Team == "Geelong") %>% 
  mutate(Date = ymd(Date)) %>% 
  select(-Venue, -Status) %>% 
  right_join(filter(results_bye, Team == "Geelong")) %>% 
  mutate(Diff = ifelse(Status == "Home.Team", Home.Odds - Away.Odds, Away.Odds - Home.Odds), 
         Expected = ifelse(Diff < 0, "W", "L")) %>% 
  count(bye, Result, Expected) %>% 
  pander(split.table = Inf,
         justify = c("left", "left", "left", "right"))
```

# Conclusion
Geelong's win/loss record after the bye does seem to be somewhat unusual. Losses after the bye compared with before are significant and unexpected.

Six of the eight post-bye losses have been away games, three of which were against teams with strong records at home. Scheduling might therefore explain some of the results.
