---
title: 'AFL: post-bye results'
author: "Neil Saunders"
date: "`r Sys.time()`"
output:
  github_document:
    toc: true
    fig_width: 9
    fig_height: 6
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE)

library(tidyverse)
library(readxl)
library(fitzRoy)
library(knitr)
library(kableExtra)
library(broom)

theme_set(theme_bw())

options(knitr.kable.NA = "")

# current year
current_year = year(Sys.Date())
```

# Summary
Since the introduction in 2012 of mid-season byes to the AFL competition, there has been a lot of discussion in the media and online regarding the so-called "curse of the bye": a perceived tendency to lose the game following a bye round.

A lot of this discussion uses simplistic counts of wins and losses for each team in post-bye games. However, this fails to take three important factors into account.

1. Whether one or both teams in the game came off a bye round
1. Scheduling - whether the post-bye team played at home or away
1. Expected result - whether the post-bye team was the favourite to win

When these variables are examined, we see that in general only a small number of post-bye games each season have an unexpected result.


# Data
Our data comes from 3 sources:

- the fitzRoy package for match dates, rounds and results
- a custom CSV file `afl_byes.csv` containing data about bye rounds derived from AFL Tables
- historical betting odds data from [Historical AFL Results and Odds Data
](https://www.aussportsbetting.com/data/historical-afl-results-and-odds-data/)

# Processing

- fetch results from AFL Tables data for season 2011 onwards.
- read the CSV file `afl_byes.csv`
- read the `afl.xlsx` file with historical odds

Then:

- pivot the odds data to long form, rename teams and flag the favourites (lowest odds)
- pivot the results data, rename teams, join with the odds and byes data, add margins and results
- join with pivoted results data to add the opponent team

```{r fetch-results}
afldata <- fetch_results_afltables(season = 2011:current_year)
```

```{r read-byes-data}
afl_byes <- read_csv("../../../../data/afl_byes.csv")
```

```{r read-odds-data}
url <- "https://www.aussportsbetting.com/historical_data/afl.xlsx"
destfile <- "/tmp/afl.xlsx"
# if(!file.exists(destfile))
curl::curl_download(url, destfile)

afl_odds <- read_excel(destfile, skip = 1)
```

```{r pivot-odds-data}
afl_odds_long <- afl_odds %>% 
  select(Date, Venue, `Home Team`, `Away Team`, `Home Odds`, `Away Odds`) %>% 
  mutate(Favourite = case_when(
    `Home Odds` < `Away Odds` ~ `Home Team`, 
    `Away Odds` < `Home Odds` ~ `Away Team`, 
    TRUE ~ NA_character_)) %>% 
  pivot_longer(3:4, names_to = "HA.Name", values_to = "Team") %>% 
  mutate(Odds = ifelse(HA.Name == "Home Team", `Home Odds`, `Away Odds`),
         Favourite = case_when(
           Team == Favourite ~ "Y",
           Team != Favourite ~ "N",
           TRUE ~ NA_character_),
         Date = ymd(Date), 
         Team = recode(Team, Brisbane = "Brisbane Lions", `GWS Giants` = "GWS")) %>% 
  select(Date, Venue, Team, Odds, Favourite)
```

```{r create-byes-dataset}
games_teams <- afldata %>% 
  select(Game, Home.Team, Away.Team) %>% 
  pivot_longer(2:3, 
               names_to = "HA.Name", 
               values_to = "Team") %>% 
  mutate(Team = recode(Team, Footscray = "Western Bulldogs"))

post_bye_games <- afldata %>% 
  select(Game, Season, Date, Round, Round.Number, Home.Team, Away.Team, Home.Points, Away.Points) %>%
  pivot_longer(6:7, names_to = "HA.Name", values_to = "Team") %>% 
  mutate(Team = ifelse(Team == "Footscray", "Western Bulldogs", Team)) %>% 
  left_join(afl_byes, by = c("Season", "Round" = "NextRound", "Team")) %>% 
  mutate(Margin = ifelse(HA.Name == "Home.Team", Home.Points - Away.Points, Away.Points - Home.Points)) %>%
  filter(!is.na(Round.y)) %>% 
  rename(ByeRound = Round.y) %>% 
  mutate(Result = case_when(
    Margin > 0 ~ "W", 
    Margin < 0 ~ "L", 
    TRUE ~ "D")) %>% 
  left_join(afl_odds_long) %>% 
  left_join(games_teams, 
            by = "Game",
            relationship = "many-to-many") %>% 
  filter(Team.x != Team.y) %>% 
  rename(HA.Name = HA.Name.x, 
         Team = Team.x, 
         Opponent = Team.y) %>% 
  select(-HA.Name.y)
```

The final dataset `post_bye_games` looks like this. It gets written to CSV.

```{r glimpse-byes-dataset-csv}
post_bye_games %>% 
  glimpse(width = 160)

post_bye_games %>% 
  write_csv("../../../../data/post_bye_games.csv")
```

# Analysis

## Visualization of bye rounds

### Bye rounds by season

In 2011 there was a bye every round, featuring at least one team.

From 2012 onwards the mid-season bye was introduced. Most commonly there are 3 consecutive bye rounds with 6 teams per round.

However, there are variations on the number of bye rounds, teams per round and which rounds are byes.

Note that 2020 was a disrupted season due to COVID. For example, the R3 game was postponed due to a player testing positive.

```{r bye-rounds-by-season}
seasons_rounds <- expand.grid(Season = 2011:current_year,
                              Round = paste0("R", 1:24))

afl_byes %>% 
  count(Season, Round) %>%
  right_join(seasons_rounds) %>% 
  mutate(Round = as.numeric(gsub("R", "", Round))) %>% 
  ggplot(aes(Round, Season)) + 
  geom_tile(aes(fill = n), color = "black") + 
  scale_fill_viridis_c(name = "Teams", na.value = "white") + 
  scale_y_continuous(breaks = 2011:current_year) +
  coord_equal() +
  theme(legend.position = "top",
        panel.grid = element_blank()) +
  labs(title = "Number of teams on a bye by season and round",
       subtitle = paste0("2011-", current_year))
```

### Bye rounds by team

We see the 2011 byes scattered through each round, and more concentrated values for the mid-season byes.

Notably Collingwood's bye has fallen in Round 13 six times.

```{r bye-rounds-by-team}
teams_rounds <- expand.grid(Team = unique(post_bye_games$Team),
                            Round = paste0("R", 1:24))

afl_byes %>% 
  count(Team, Round) %>% 
  right_join(teams_rounds) %>% 
  mutate(Round = as.numeric(gsub("R", "", Round))) %>% 
  ggplot(aes(Round, Team)) + 
  geom_tile(aes(fill = n), color = "black") + 
  scale_fill_viridis_c(name = "Count", na.value = "white") + 
  scale_y_discrete(limits = rev) +
  coord_equal() +
  theme(legend.position = "top",
        panel.grid = element_blank()) +
  labs(title = "Byes per round by team",
       subtitle = paste0("2011-", current_year))
```

### Bye rounds by season and team

Another attempt to visualise the bye rounds. Again we see the difference between the 2011 byes and the mid-season byes.

It's interesting that we see contiguous blocks - is the draw for byes as much alphabetical sometimes as anything else!

```{r bye-rounds-by-season-team}
afl_byes %>% 
  count(Season, Team, Round) %>% 
  mutate(Round = as.numeric(gsub("R", "", Round))) %>% 
  ggplot(aes(Round, Team)) + 
  geom_tile(aes(fill = n), color = "black") + 
  scale_fill_viridis_c(na.value = "white") +
  scale_y_discrete(limits = rev) +
  facet_wrap(~Season, ncol = 5) + 
  guides(fill = "none") + 
  theme(axis.text.y = element_text(size = 6),
        legend.position = "top",
        panel.grid = element_blank()) +
  labs(title = "Team byes by season and round",
       subtitle = "2011-2023")
```

### Post-bye rounds by teams coming off the bye

Post-bye ames might feature one or both teams coming off a bye. If the bye were a factor influencing the post-bye game, we might assume the effect would be stronger when only one team comes off a bye.

In most seasons (disregarding 2011) there are 2-5 games where both teams come off a bye, but the majority involve one post-bye team.

```{r post-bye-one-both}
post_bye_games %>% 
  group_by(Season, Game) %>% 
  summarise(N = n()) %>% 
  ungroup() %>% 
  count(Season, N) %>% 
  mutate(`teams off bye` = ifelse(N == 1, "one", "two")) %>% 
  ggplot(aes(Season, n, fill = `teams off bye`)) + 
  geom_col() + 
  geom_text(aes(label = n), 
            color = "white",
            position = position_stack(vjust = 0.5)) +
  labs(y = "Count",
       title = "Count of post-bye games by season and number of teams coming off bye",
       subtitle = paste0("2011-", current_year)) +
  scale_fill_brewer(palette = "Set1") +
  scale_x_continuous(breaks = seq(2011, current_year, 2)) +
  theme(legend.position = "top")
```

## Wins and losses

Let's start looking at wins and losses in post-bye games. In the following sections we look only at games where one team comes off a bye. We include season 2011 even though byes in that season were not mid-season.

### Distribution of margins by expected result

One important factor missing from many analyses looking at post-bye games is: what was the expected result? For the purposes of this report, expected result means "who was favourite according to the betting odds".

Plotting post-bye game margins versus expected winner shows that on average, favourites win and non-favourites lose. But there are always exceptions!


```{r post-bye-margin-odds}
post_bye_games %>% 
  group_by(Game) %>% 
  filter(n() == 1) %>% 
  ungroup() %>% 
  mutate(Favourite = recode(Favourite, Y = "Yes", N = "No")) %>% 
  ggplot(aes(fct_relevel(Favourite, "Yes", "No"), Margin)) + 
  geom_jitter(width = 0.2) +
  geom_boxplot(aes(fill = Favourite),
               outlier.shape = NA,
               alpha = 0.4) + 
  geom_hline(yintercept = 0,
             linetype = "dotted") +
  scale_fill_brewer(palette = "Set1") +
  guides(fill = "none") +
  labs(x = "Favourite",
       y = "Margin",
       title = "Margin in post-bye games by expected result",
       subtitle = paste0("2011-", current_year))
```

### Overall

A simple count of post-bye results shows that more teams lose coming off a bye then win.

However, this doesn't tell us very much about why this should be the case.

```{r post-bye-win-loss}
post_bye_games %>% 
  group_by(Game) %>% 
  filter(n() == 1) %>% 
  ungroup() %>% 
  count(Result) %>% 
  mutate(Result = recode(Result, W = "Win", L = "Loss")) %>% 
  ggplot(aes(fct_relevel(Result, "Win", "Loss"), n)) + 
  geom_col(fill = "#377EB8") + 
  geom_text(aes(label = n), 
            color = "white",
            vjust = 5) +
  labs(x = "Result", 
       y = "Count", 
       title = "Results for AFL teams coming off a bye", 
       subtitle = paste0("2011-", current_year))
```

### Overall by home/away

Scheduling plays an important role in influencing post-bye results.

More teams coming off a bye were playing away from home. And away teams are more likely to lose a game.


```{r post-bye-home-away}
post_bye_games %>% 
  group_by(Game) %>% 
  filter(n() == 1) %>% 
  ungroup() %>% 
  count(HA.Name, Result) %>% 
  mutate(Result = recode(Result, W = "Win", L = "Loss"),
         Result = fct_relevel(Result, "Win", "Loss"),
         HA.Name = fct_relevel(HA.Name, "Home.Team", "Away.Team")) %>% 
  ggplot(aes(HA.Name, n, fill = Result)) + 
  geom_col() + 
  geom_text(aes(label = n), 
            color = "white",
            position = position_stack(vjust = 0.5)) +
  labs(x = "Home/Away", 
       y = "Count", 
       title = "Results for AFL teams coming off a bye by home/away", 
       subtitle = paste0("2011-", current_year)) +
  scale_fill_brewer(palette = "Set1") +
  theme(legend.position = "top")
```

The home/away win/loss difference has p < 0.05 in a Fisher's exact test.

```{r post-bye-home-away-fisher}
post_bye_games %>% 
  group_by(Game) %>% 
  filter(n() == 1) %>% 
  ungroup() %>% 
  count(HA.Name, Result) %>% 
  mutate(Result = recode(Result, W = "Win", L = "Loss"),
         HA.Name = fct_relevel(HA.Name, "Home.Team", "Away.Team")) %>% 
  pivot_wider(names_from = "HA.Name", values_from = "n") %>% 
  kbl() %>% 
  kable_styling(bootstrap_options = c("striped", "compact"))

ftest <- post_bye_games %>% 
  group_by(Game) %>% 
  filter(n() == 1) %>% 
  ungroup() %>% 
  count(HA.Name, Result) %>% 
  mutate(Result = recode(Result, W = "Win", L = "Loss"),
         HA.Name = fct_relevel(HA.Name, "Home.Team", "Away.Team")) %>% 
  pivot_wider(names_from = "HA.Name", 
              values_from = "n") %>% 
  column_to_rownames("Result") %>% 
  fisher.test()

ftest %>% 
  tidy() %>% 
  kbl() %>% 
  kable_styling(bootstrap_options = c("striped", "compact"))
```

### Overall by expected result

When we look at post-bye results by expected result, it's clear that a majority of favourites win, and a majority of non-favourites lose.

```{r post-bye-expected-result}
post_bye_games %>% 
  group_by(Game) %>% 
  filter(n() == 1) %>% 
  ungroup() %>% 
  count(Favourite, Result) %>% 
  mutate(Result = recode(Result, W = "Win", L = "Loss"),
         Result = fct_relevel(Result, "Win", "Loss"),
         Favourite = recode(Favourite, Y = "Yes", N = "No"),
         Favourite = fct_relevel(Favourite, "Yes", "No")) %>% 
  ggplot(aes(Favourite, n, fill = Result)) + 
  geom_col() + 
  geom_text(aes(label = n), 
            color = "white",
            position = position_stack(vjust = 0.5)) +
  labs(x = "Favourite", 
       y = "Count", 
       title = "Results for AFL teams coming off a bye by expected result", 
       subtitle = paste0("2011-", current_year)) +
  scale_fill_brewer(palette = "Set1") +
  theme(legend.position = "top")
```

### By season

If coming off a bye were a consistent disadvantage, we might expect to see post-bye teams losing at a similar rate across seasons.

In fact we see all combinations: seasons where teams in post-bye games have more losses, more wins and equal numbers of losses and wins.

```{r post-bye-season}
post_bye_games %>% 
  group_by(Game) %>% 
  filter(n() == 1) %>% 
  ungroup() %>% 
  count(Season, Result) %>% 
  mutate(Result = recode(Result, W = "Win", L = "Loss")) %>% 
  ggplot(aes(fct_relevel(Result, "Win", "Loss"), n)) + 
  geom_col(fill = "#377EB8") + 
  labs(x = "Result", 
       y = "Count", 
       title = "Results for AFL teams coming off a bye by season", 
       subtitle = paste0("2011-", current_year)) +
  facet_wrap(~Season, ncol = 5)
```

### By season and expected result

When we include expected results we see that in a majority of cases, favourites win post-bye games and non-favourites lose them.

```{r post-bye-season-expected-result}
post_bye_games %>% 
  group_by(Game) %>% 
  filter(n() == 1) %>% 
  ungroup() %>% 
  count(Season, Favourite, Result) %>% 
  mutate(Result = recode(Result, W = "Win", L = "Loss"),
         Result = fct_relevel(Result, "Win", "Loss"),
         Favourite = recode(Favourite, Y = "Yes", N = "No"),
         Favourite = fct_relevel(Favourite, "Yes", "No")) %>% 
  ggplot(aes(Favourite, n)) + 
  geom_col(aes(fill = Result)) + 
  labs(x = "Favourite", 
       y = "Count", 
       title = "Results for AFL teams coming off a bye by season and expected result", 
       subtitle = paste0("2011-", current_year)) +
  scale_fill_brewer(palette = "Set1") +
  facet_wrap(~Season, ncol = 5) +
  theme(legend.position = "top")
```

### By team

Do some teams suffer from "the curse of the bye"?

Certainly, using a simple win-loss count, some teams do seem to lose a lot of post-bye games: notably Geelong and Gold Coast.

```{r post-bye-team}
post_bye_games %>% 
  group_by(Game) %>% 
  filter(n() == 1) %>% 
  ungroup() %>% 
  count(Team, Result) %>% 
  mutate(Result = recode(Result, W = "Win", L = "Loss")) %>% 
  ggplot(aes(fct_relevel(Result, "Win", "Loss"), n)) + 
  geom_col(fill = "#377EB8") + 
  labs(x = "Result", 
       y = "Count", 
       title = "Results for AFL teams coming off a bye by team", 
       subtitle = paste0("2011-", current_year)) +
  facet_wrap(~Team, ncol = 6)
```

### By team and expected result

However, once again the introduction of expected result can paint a different picture.

Gold Coast, for example, were not the favourite to win in any of their post-bye losses. In other words they lose a lot of games in general, not just post-bye games.

Geelong on the other hand were favourites in the majority of their post-bye losses.

```{r post-bye-team-expected-result}
post_bye_games %>% 
  group_by(Game) %>% 
  filter(n() == 1) %>% 
  ungroup() %>% 
  count(Team, Favourite, Result) %>% 
  mutate(Result = recode(Result, W = "Win", L = "Loss"),
         Result = fct_relevel(Result, "Win", "Loss"),
         Favourite = recode(Favourite, Y = "Yes", N = "No"),
         Favourite = fct_relevel(Favourite, "Yes", "No")) %>% 
  ggplot(aes(Favourite, n)) + 
  geom_col(aes(fill = Result)) + 
  labs(x = "Favourite", 
       y = "Count", 
       title = "Results for AFL teams coming off a bye by team and expected result", 
       subtitle = paste0("2011-", current_year)) +
  scale_fill_brewer(palette = "Set1") +
  facet_wrap(~Team, ncol = 6) +
  theme(legend.position = "top")
```

### Unexpected wins and losses by season

Here, we try a few ways to visualise post-bye results by season and expectation.

First: combine expectation and result into a new variable, then plot the proportion of unexpected results by season.

```{r post-bye-unexpected-01}
post_bye_games %>% 
  group_by(Game) %>% 
  filter(n() == 1) %>% 
  ungroup() %>% 
  count(Season, Favourite, Result) %>% 
  mutate(Status = paste0(Favourite, Result)) %>% 
  mutate(Status = case_when(
    Status == "NL" ~ "Expected loss", 
    Status == "NW" ~ "Unexpected win", 
    Status == "YL" ~ "Unexpected loss", 
    Status == "YW" ~ "Expected win")) %>% 
  ggplot(aes(Season, n)) + 
  geom_col(aes(fill = Status)) + 
  scale_fill_brewer(palette = "OrRd") +
  scale_x_continuous(breaks = seq(2011, current_year, 2)) +
  labs(y = "Count",
       title = "Post-bye results by season and expectation",
       subtitle = paste0("2011-", current_year)) +
  theme(legend.position = "top")
```

Second - a similar idea but combine expected wins and losses into one "expected result" variable, to show more clearly unexpected wins and losses as a proportion.

```{r post-bye-unexpected-02}
post_bye_games %>% 
  group_by(Game) %>% 
  filter(n() == 1) %>% 
  ungroup() %>% 
  count(Season, Favourite, Result) %>% 
  mutate(Status = paste0(Favourite, Result)) %>% 
  mutate(Status = case_when(
    Status == "NL" ~ "Expected result", 
    Status == "NW" ~ "Unexpected win", 
    Status == "YL" ~ "Unexpected loss", 
    Status == "YW" ~ "Expected result")) %>% 
  ggplot(aes(Season, n)) + 
  geom_col(aes(fill = Status)) + 
  scale_fill_brewer(palette = "OrRd") +
  scale_x_continuous(breaks = seq(2011, current_year, 2)) +
  labs(y = "Count",
       title = "Post-bye results by season and expectation",
       subtitle = paste0("2011-", current_year)) +
  theme(legend.position = "top")
```

Third, split the data by expectation and plot the proportion of wins and losses.

```{r post-bye-unexpected-03}
post_bye_games %>% 
  group_by(Game) %>% 
  filter(n() == 1) %>% 
  ungroup() %>% 
  count(Season, Favourite, Result) %>% 
  mutate(Result = recode(Result, W = "Win", L = "Loss"),
         Result = fct_relevel(Result, "Win", "Loss"),
         Favourite = recode(Favourite, Y = "Yes", N = "No"),
         Favourite = fct_relevel(Favourite, "Yes", "No")) %>% 
  ggplot(aes(Season, n)) + 
  geom_col(aes(fill = Result)) +
  facet_grid(Favourite ~ .) +
  labs(y = "Count",
       title = "Post-bye results by season and expectation",
       subtitle = paste0("2011-", current_year)) +
  scale_fill_brewer(palette = "Set1") +
  scale_x_continuous(breaks = seq(2011, current_year, 2)) +
  theme(legend.position = "top")
```

Fourth, filter for favourite/loss or non-favourite/win and just plot the counts of unexpected wins and losses.

This one shows that unexpected losses are slightly more frequent than unexpected wins.

```{r post-bye-unexpected-04}
post_bye_games %>% 
  group_by(Game) %>% 
  filter(n() == 1) %>% 
  ungroup() %>% 
  count(Season, Favourite, Result) %>% 
  mutate(Expectation = paste0(Favourite, Result),
         Result = recode(Result, W = "Win", L = "Loss"),
         Result = fct_relevel(Result, "Win", "Loss")) %>% 
  filter(Expectation %in% c("NW", "YL")) %>%
  ggplot(aes(Season, n, fill = Result)) + 
  geom_col() + 
  facet_wrap(~Result, ncol = 1) +
  labs(y = "Count",
       title = "Unexpected post-bye results by season",
       subtitle = paste0("2011-", current_year)) +
  scale_fill_brewer(palette = "Set1") +
  scale_x_continuous(breaks = seq(2011, current_year, 2)) +
  guides(fill = "none")
```

## Comparison of expected results with other games

```{r calculate-special}
# do some calculations here because inline backtick issues

games_drawn <- afldata %>% 
  filter(Margin == 0)

games_no_favourite <- afl_odds %>% 
  filter(`Home Odds` == `Away Odds`)
```

This section is somewhat experimental.

We might ask: is the proportion of post-bye games with unexpected results very different to other games?

The question then is: with which "other games" should we compare? Some ideas are:

- all of the games that were not post-bye over the same period
- all of the games that were not post-bye in the same rounds
- all of the games that were not post-bye on the same dates

To address this, we generate 3 new datasets corresponding to each set in the list, and combine Favourite and Result into a new variable. Then we can count expected results (favourite wins or non-favourite loses) and unexpected results (favourite loses, non-favourite wins).

Some notes:

In these datasets games are from the perspective of the home team (clearly if one team loses the other wins and if one is the favourite, the other is not, so we only need one row per game).

Drawn games are omitted. There are `r nrow(games_drawn)` drawn games in this time period.

Games with no favourite (equal odds) are omitted. There are `r nrow(games_no_favourite)` such games in this time period.

**Summary** - we see:

- a slightly _higher_ percentage of unexpected results in post-bye games compared with non-post-bye games played in the same rounds, or on the same dates
- a slightly _lower_ percentage of unexpected results in post-bye games compared with all non-post-bye games in the same time period

It's probably best not to over-interpret here, until we have a clearer idea about what makes a good comparison.

```{r count-expected}
# all non-post-bye games
non_post_bye_games_all <- afldata %>% 
  filter(!Game %in% post_bye_games$Game) %>% 
  select(Game, Season, Date, Round, Round.Number, Home.Team, Away.Team, Home.Points, Away.Points) %>%
  pivot_longer(6:7, 
               names_to = "HA.Name", 
               values_to = "Team") %>% 
  mutate(Team = ifelse(Team == "Footscray", "Western Bulldogs", Team),
         Margin = ifelse(HA.Name == "Home.Team", Home.Points - Away.Points, Away.Points - Home.Points),
         Result = case_when(
           Margin > 0 ~ "W", 
           Margin < 0 ~ "L", 
           TRUE ~ "D")) %>% 
  left_join(afl_odds_long) %>% 
  mutate(Expected = paste0(Favourite, Result)) %>% 
  filter(HA.Name == "Home.Team")

# non-post-bye games in the same rounds:
non_post_bye_games_round <- afldata %>% 
  filter(paste0(Season, Round) %in% paste0(post_bye_games$Season, post_bye_games$Round), 
         !Game %in% post_bye_games$Game) %>% 
  select(Game, Season, Date, Round, Round.Number, Home.Team, Away.Team, Home.Points, Away.Points) %>%
  pivot_longer(6:7, 
               names_to = "HA.Name", 
               values_to = "Team") %>% 
  mutate(Team = ifelse(Team == "Footscray", "Western Bulldogs", Team),
         Margin = ifelse(HA.Name == "Home.Team", Home.Points - Away.Points, Away.Points - Home.Points),
         Result = case_when(
           Margin > 0 ~ "W", 
           Margin < 0 ~ "L", 
           TRUE ~ "D")) %>% 
  left_join(afl_odds_long) %>% 
  mutate(Expected = paste0(Favourite, Result)) %>% 
  filter(HA.Name == "Home.Team")

# non-post-bye games on the same date:
non_post_bye_games_date <- afldata %>% 
filter(Date %in% unique(post_bye_games$Date), 
       !Game %in% post_bye_games$Game) %>% 
  select(Game, Season, Date, Round, Round.Number, Home.Team, Away.Team, Home.Points, Away.Points) %>%
  pivot_longer(6:7, 
               names_to = "HA.Name", 
               values_to = "Team") %>% 
  mutate(Team = ifelse(Team == "Footscray", "Western Bulldogs", Team),
         Margin = ifelse(HA.Name == "Home.Team", Home.Points - Away.Points, Away.Points - Home.Points),
         Result = case_when(
           Margin > 0 ~ "W", 
           Margin < 0 ~ "L", 
           TRUE ~ "D")) %>% 
  left_join(afl_odds_long) %>% 
  mutate(Expected = paste0(Favourite, Result)) %>% 
  filter(HA.Name == "Home.Team")

# now count Expected and combine with post-bye games
expected_counts <- list(
  `Post-bye games 2011-2023` = post_bye_games %>% 
    group_by(Game) %>% 
    filter(n() == 1) %>% 
    ungroup() %>% 
    mutate(Expected = paste0(Favourite, Result)) %>% 
    count(Expected),
  `All non-post-bye games 2011-2023` = non_post_bye_games_all %>% 
    count(Expected),
  `Same round non-post-bye games 2011-2023` = non_post_bye_games_round %>% 
    count(Expected),
  `Same date non-post-bye games 2011-2023` = non_post_bye_games_date %>% 
    count(Expected)) %>% 
  bind_rows(.id = "dataset")

# tabulate
expected_counts %>% 
  filter(Expected %in% c("YW", "YL", "NW", "NL")) %>% 
  mutate(Expected = recode(Expected, 
                           NL = "No/Loss", 
                           NW = "No/Win", 
                           YL = "Yes/Loss", 
                           YW = "Yes/Win")) %>% 
  pivot_wider(names_from = "Expected", 
              values_from = "n") %>% 
  select(dataset, `Yes/Win`, `No/Loss`, `Yes/Loss`, `No/Win`) %>% 
  mutate(`% unexpected` = 100 * (`Yes/Loss` + `No/Win`) / (`Yes/Win` + `No/Loss` + `Yes/Loss` + `No/Win`)) %>% 
  kbl(caption = "Game results by favourite 2011-2023", digits = 2) %>%  
  add_header_above(c(" ", "Favourite/Result" = 4, " "))
```



## Comparison to media articles

Now let's look at some media articles about post-bye results and compare.

### SEN June 25 2023

[The numbers that don't bode well for the weekend’s bye teams](https://www.sen.com.au/news/2023/06/25/the-numbers-that-dont-bode-well-for-the-weekends-bye-teams/)

>Who is a fan of the bye?
So far this year, teams coming off the bye own a pitiful 2-10 win-loss record from 12 matches.
Of those dozen matches, only two have featured both teams coming off the bye - Sydney v St Kilda in Round 13 and Collingwood v Adelaide in Round 15.
Let’s take those results out of it and it’s 0-8 for teams coming off the bye playing a non-bye team that week.

All of that is quite correct. However, when we look at the eight post-bye losses to R14, only two of the teams were expected to win, and one of those teams was playing away.

The bye rounds for 2023 have since completed: of 14 teams coming off the bye, there were ten losses and four wins. The final tally is two unexpected losses and one unexpected win.

```{r compare-media-01}
post_bye_games %>% 
  filter(Season == 2023) %>% 
  group_by(Game) %>% 
  filter(n() == 1) %>% 
  ungroup() %>% 
  select(Season, Round, Team, HA.Name, Opponent, Margin, Result, Odds, Favourite) %>%  
  kbl(caption = "Season 2023 post-bye games") %>% 
  kable_styling(bootstrap_options = c("striped", "compact"))
```

### AFL June 6 2023

[Beware the bye: How your club fares after its mid-season break](https://www.afl.com.au/news/944914/beware-the-bye-how-your-club-fares-after-its-mid-season-break)

>EVERY team should beware their mid-season bye.
Statistics show that over the past 10 years not one club has mastered how to come out the other side of their week off with consistent success.
In fact, just four clubs hold a winning record over that period in their first match after the break, with Fremantle, Hawthorn, Richmond and St Kilda all sporting a 6-4 win-loss mark since 2013.

The AFL article looks at post-bye win-loss records 2013-2022.

We can mostly reproduce their numbers by counting results by team - but they have included games where both teams came off the bye! They also have Sydney at 5-5 (wins-losses), whereas my data has them at 4-6.

Using only games with one team coming off the bye for years 2013-2022 makes quite a difference to the numbers. Brisbane for example move from 5-5 (wins-losses) to 5-1.

```{r compare-media-02}
post_bye_games_01 <- post_bye_games %>% 
  filter(between(Season, 2013, 2022)) %>% 
  count(Team, Result) %>% 
  pivot_wider(names_from = "Result", 
              values_from = "n") %>% 
  select(Team, W, D, L)

post_bye_games_02 <- post_bye_games %>% 
  filter(between(Season, 2013, 2022)) %>% 
  group_by(Game) %>% 
  filter(n() == 1) %>% 
  ungroup() %>% 
  count(Team, Result) %>% 
  pivot_wider(names_from = "Result", 
              values_from = "n") %>% 
  select(Team, W, L)

post_bye_games_01 %>% 
  left_join(post_bye_games_02,
            by = "Team") %>% 
  kbl(col.names = c("Team", "Win", "Draw", "Loss", "Win", "Loss"),
      caption = "Post-bye win-loss count 2011-2022") %>% 
  kable_styling(bootstrap_options = c("striped", "compact")) %>% 
  add_header_above(c(" ", "Count (SMH)" = 3, "Count (1 team off bye)" = 2))
```

### The Age 

The Age [Beveridge’s Bulldogs have eye on high-flying Pies after bucking bye curse](https://www.theage.com.au/sport/afl/western-bulldogs-buck-bye-curse-with-a-bang-against-slow-firing-fremantle-dockers-20230701-p5dl00.html)

It's not clear what the "bye curse" refers to in this article, since it makes no mention of the Western Bulldogs post-bye record.

However, it does say:

>Western Bulldogs coach Luke Beveridge has credited the mindset of his team to bounce back from the bye and buck the mid-season trend of teams losing upon their return

So let's focus on only mid-season byes (2012-2023), where the Bulldogs were the only team coming off the bye.

Their record is 5-4 (wins-losses) and they were not favourites to win in any of the four losses.

```{r compare-media-03}
post_bye_games %>% 
  group_by(Game) %>% 
  filter(n() == 1) %>% 
  ungroup() %>% 
  filter(between(Season, 2012, 2023), 
         Team == "Western Bulldogs") %>% 
  select(Season, Round, Team, HA.Name, Opponent, Margin, Result, Odds, Favourite) %>%  
  kbl(caption = "Post-bye result Western Bulldogs 2012-2023") %>% 
  kable_styling(bootstrap_options = c("striped", "compact"))
```

