---
title: "All-Australian Team Statistics"
author: "Neil Saunders"
date: "\\\ncompiled `r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output: 
  github_document:
    fig_width: 9
    fig_height: 6
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      fig.path = "figures/")

library(tidyverse)
library(knitr)
library(kableExtra)

theme_set(theme_bw())
```

# Summary
A look at some of the numbers around All-Australian (AA) AFL teams.

# Data
We start from the CSV file `all_australian_players.csv`. See the Appendix for how it was generated.

We do some additional processing to extract the team name and unique player name from the `href`.

Ages are calculated at the end of the year in which AA status was awarded.

```{r read-data}
all_australian_players <- read_csv("../../../../data/all_australian_players.csv")

all_australian_players <- all_australian_players %>% 
  mutate(team = str_match(href, "pp-(.*?)--")[, 2],
         player = str_match(href, "--(.*?)$")[, 2])

all_australian_players %>% 
  glimpse(width = 160)
```

# Analysis

## Most Awards

### By team

This count includes players awarded AA for the team in multiple years.

```{r count-team}
all_australian_players %>% 
  count(team, sort = TRUE) %>% 
  mutate(team = str_to_title(team) %>% str_replace_all(., "-", " ")) %>% 
  ggplot(aes(reorder(team, n), n)) +
  geom_col(fill = "skyblue4") +
  coord_flip() +
  labs(x = "Team",
       y = "Count",
       title = "Total AA players by team",
       subtitle = "includes multiple award winners")
```

This count only counts one AA award per player per team.

```{r count-distinct-team}
all_australian_players %>%
  distinct(team, player) %>% 
  count(team, sort = TRUE) %>% 
  mutate(team = str_to_title(team) %>% str_replace_all(., "-", " ")) %>% 
  ggplot(aes(reorder(team, n), n)) +
  geom_col(fill = "skyblue4") +
  coord_flip() +
  labs(x = "Team",
       y = "Count",
       title = "Total AA players by team",
       subtitle = "distinct player and team")
```

### By team and season

Geelong's dominance in the late 2000s is apparent.

```{r count-team-season}
all_australian_players %>% 
  count(season, team, sort = TRUE) %>% 
  mutate(team = str_to_title(team) %>% 
           str_replace_all(., "-", " ")) %>% 
  ggplot(aes(season, fct_relevel(team, rev))) +
  geom_tile(aes(fill = n)) +
  coord_equal() +
  scale_fill_viridis_c(breaks = seq(1,9,2)) +
  labs(x = "Season",
       y = "Team",
       title = "AA player count by season and team")
```

### By player

Players with 5+ AA awards.

```{r count-player}
all_australian_players %>% 
  count(player, sort = TRUE) %>% 
  filter(n > 4) %>% 
  mutate(player = str_to_title(player) %>% str_replace_all(., "-", " ")) %>% 
  ggplot(aes(reorder(player, n), n)) +
  geom_col(fill = "skyblue4") +
  coord_flip() +
  labs(x = "Count",
       y = "Player",
       title = "AA awards by player",
       subtitle = "Players with 5+ awards")
```

### By player and team

Players with 5+ AA awards for one team. This drops some prominent players _e.g._ Gary Ablett Jnr. and Lance Franklin, who each had 4 awards at 2 different clubs for 8 total. 

```{r count-player-team}
all_australian_players %>% 
  count(player, team, sort = TRUE) %>% 
  filter(n > 4) %>% 
  mutate(player = str_to_title(player) %>% str_replace_all(., "-", " ")) %>% 
  ggplot(aes(reorder(player, n), n)) +
  geom_col(fill = "skyblue4") +
  coord_flip() +
  labs(x = "Count",
       y = "Player",
       title = "AA awards by player per team",
       subtitle = "Players with 5+ awards for one or more teams")
```


## Player selection

### Youngest

The top 20 youngest AA players.

```{r sort-age-youngest}
all_australian_players %>% 
  group_by(player) %>% 
  filter(age_year_end_days == min(age_year_end_days)) %>% 
  ungroup() %>% 
  arrange(age_year_end_days) %>% 
  slice(1:20) %>% 
  select(-c(href, player)) %>% 
  mutate(team = str_to_title(team) %>% str_replace_all(., "-", " ")) %>% 
  kbl(caption = "20 youngest AA players by year first selected",
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

### Youngest by team

The all-time youngest AA player by team.

```{r sort-age-team-youngest}
all_australian_players %>% 
  group_by(team) %>% 
  filter(age_year_end_days == min(age_year_end_days)) %>% 
  ungroup() %>% 
  select(-c(href, player)) %>% 
  mutate(team = str_to_title(team) %>% str_replace_all(., "-", " ")) %>%
  arrange(team) %>% 
  kbl(caption = "Youngest AA player by team",
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

### Youngest by season

The youngest AA player each season.

```{r sort-age-season-youngest}
all_australian_players %>% 
  group_by(season) %>% 
  filter(age_year_end_days == min(age_year_end_days)) %>% 
  ungroup() %>% 
  select(-c(href, player)) %>% 
  mutate(team = str_to_title(team) %>% str_replace_all(., "-", " ")) %>%
  arrange(season) %>% 
  kbl(caption = "Youngest AA player by season",
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))
```


### Oldest

The top 20 oldest AA players.

```{r sort-age-oldest}
all_australian_players %>% 
  group_by(player) %>% 
  filter(age_year_end_days == max(age_year_end_days)) %>% 
  ungroup() %>% 
  arrange(desc(age_year_end_days)) %>% 
  slice(1:20) %>% 
  select(-c(href, player)) %>% 
  mutate(team = str_to_title(team) %>% str_replace_all(., "-", " ")) %>% 
  kbl(caption = "20 oldest AA players by year last selected",
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

### Oldest by team

The all-time oldest AA player by team.

```{r sort-age-team-oldest}
all_australian_players %>% 
  group_by(team) %>% 
  filter(age_year_end_days == max(age_year_end_days)) %>% 
  ungroup() %>% 
  select(-c(href, player)) %>% 
  mutate(team = str_to_title(team) %>% str_replace_all(., "-", " ")) %>%
  arrange(team) %>% 
  kbl(caption = "Oldest AA player by team",
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))
```


### Oldest by season

The oldest AA player each season.

```{r sort-age-season-oldest}
all_australian_players %>% 
  group_by(season) %>% 
  filter(age_year_end_days == max(age_year_end_days)) %>% 
  ungroup() %>% 
  select(-c(href, player)) %>% 
  mutate(team = str_to_title(team) %>% str_replace_all(., "-", " ")) %>%
  arrange(season) %>% 
  kbl(caption = "Oldest AA player by season",
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))
```


## Trends

Any obvious trends in age, height or weight?

More AA recent players (2005 onwards) are taller on average, though there has been a small decline in the most recent years.

```{r plot-variables}
all_australian_players %>% 
  select(season, age_year_end_years, height, weight) %>% 
  pivot_longer(-season) %>% 
  ggplot(aes(season, value)) +
  geom_jitter(aes(group = season),
              size = 1,
              color = "grey80",
              width = 0.2) +
  stat_summary(geom = "crossbar",
               color = "red",
               fun = mean) +
  facet_grid(name ~ ., scales = "free_y")
```


```{r mean-min-variables}
all_australian_players %>% 
  group_by(season) %>% 
  summarise(mean_age = mean(age_year_end_years), 
            mean_height = mean(height), 
            mean_weight = mean(weight)) %>% 
  ungroup() %>% 
  pivot_longer(-season) %>% 
  group_by(name) %>% 
  filter(value == min(value)) %>% 
  ungroup() %>% 
  arrange(name) %>% 
  kbl(caption = "Season with the youngest, shortest and lightest AA team",
      digits = 2) %>% 
  kable_styling(bootstrap_options = c("striped", "condensed"))
```


```{r mean-max-variables}
all_australian_players %>% 
  group_by(season) %>% 
  summarise(mean_age = mean(age_year_end_years), 
            mean_height = mean(height), 
            mean_weight = mean(weight)) %>% 
  ungroup() %>% 
  pivot_longer(-season) %>% 
  group_by(name) %>% 
  filter(value == max(value)) %>% 
  ungroup() %>% 
  arrange(name) %>% 
  kbl(caption = "Season with the oldest, tallest, heaviest AA team",
      digits = 2) %>% 
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

## Sydney Swans

Special interest :)

All Swans AA players.

```{r sydney-swans}
all_australian_players %>% 
  filter(str_detect(team, "sydney-swans")) %>%
  select(-c(href, player, team)) %>% 
  arrange(season) %>% 
  kbl(caption = "All Sydney Swans AA players",
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

# Appendix

Creating the AA players dataset.

```{r echo=TRUE, eval=FALSE}
## get Footywire AA players by season
get_aa <- function(season) {
  aa <- read_html(paste0("https://www.footywire.com/afl/footy/all_australian_selection?year=", season))
  
  n <- case_when(
    season > 1997 ~ 22,
    between(season, 1994, 1997) ~ 21,
    between(season, 1991, 1993) ~ 20
  )
  
  aa_df <- tibble(
    season = season,
    href = aa %>%
      html_nodes("a") %>% 
      html_attr("href") %>% 
      .[str_detect(., "^pp")] %>% 
      .[1:n])
  
  aa_df
}

## get player Footywire data using an href
get_player_data <- function(href) {
  print(href)
  u <- paste0("https://www.footywire.com/afl/footy/", href)
  p <- read_html(u)
  
  player <- tibble(
    name = p %>% html_node("#playerProfileName") %>% html_text(),
    dob = p %>% html_node("#playerProfileData1") %>% html_text() %>% str_match(., "Born: (.*?\\d{4}) ") %>% .[, 2] %>% mdy(),
    height = p %>% html_node("#playerProfileData2") %>% html_text() %>% str_match(., "Height: (.*?)cm") %>% .[, 2] %>% as.numeric(),
    weight = p %>% html_node("#playerProfileData2") %>% html_text() %>% str_match(., "Weight: (.*?)kg") %>% .[, 2] %>% as.numeric(),
  )
  
  player
}

# start by creating a data frame with seasons and the AA team nested
# then unnest to get 2 columns: season and href (link to the players)
all_australian_players <- tibble(
  season = 1991:2022
) %>% 
  mutate(aa = map(season, get_aa)) %>% 
  select(-season) %>% 
  unnest(aa)

# then use href to get player data nested
# and unnest again
# finally calculate the ages from dob and season
# can then write to all_australian_players.csv
all_australian_players <- all_australian_players %>% 
  mutate(data = map(href, get_player_data)) %>% 
  unnest(data) %>% 
  mutate(age_year_end_days = as.numeric(ymd(paste(season, 12, 31)) - dob), 
         age_year_end_years = age_year_end_days / 365)
```
