---
title: "AFL/VFL Drawn Games"
author: "Neil Saunders"
date: "compiled `r Sys.time()`"
output: 
  html_document: 
    keep_md: yes
    number_sections: yes
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyr)
library(dplyr)
library(knitr)
library(lubridate)

games <- readLines("http://afltables.com/afl/stats/biglists/bg8.txt") %>%
  gsub("([[:alpha:]]+) +([[:alpha:]]+)","'\\1 \\2'", .) %>%
  gsub("'([SPGQE]F) (.*?)' ", "\\1  \\2", .) %>%
  gsub("Cazaly''s Stadium'", "'Cazalys Stadium'", .) %>%
  textConnection() %>%
  read.table(stringsAsFactors = FALSE, skip = 2, header = FALSE) %>%
  rename("game" = V1, "Date" = V2, "round" = V3, "team1" = V4, "score1" = V5, "team2" = V6, "score2" = V7, "venue" = V8) %>%
  mutate(Year = as.numeric(substring(Date, nchar(Date) - 3, nchar(Date))),
         team1 = gsub("([a-z])([A-Z])", "\\1 \\2", team1),
         team2 = gsub("([a-z])([A-Z])", "\\1 \\2", team2))
```

# Questions
Compiled by user *Thunder Shaker* on the [RWO Swans Fan Forum](http://www.redandwhiteonline.com/forum/showthread.php/43085-Lies-damned-lies-and-statistics?p=732150&viewfull=1#post732150).

[Published report at RPubs](http://rpubs.com/neilfws/292080).

## How many drawn games have there been?
`r games %>% nrow()`.

## Which club has been involved in the greatest number of drawn games?
```{r q2}
games %>% 
  select(team1, team2)  %>% 
  gather(team, name) %>% 
  count(name)  %>% 
  arrange(desc(n)) %>% 
  slice(1:10) %>%
  kable()
```

## Which two clubs have been involved in the most draws with each other? (Two correct answers)
```{r q3}
games %>% 
  mutate(teams = apply(games[, c("team1", "team2")], 1, function(x) paste(sort(x), collapse = "/"))) %>% 
  count(teams) %>% 
  arrange(desc(n)) %>% 
  slice(1:10) %>%
  kable()
```

## Which current club has played the most games since their last draw?
```{r q4}
games %>% 
  mutate(Date = as_date(Date, "%d-%b-%Y")) %>% 
  select(team1, team2, round, Date) %>% 
  gather(team, name, -round, -Date) %>% 
  arrange(name, Date) %>% 
  mutate(diff = as_date(Sys.Date()) - Date) %>% 
  filter(diff >= 0) %>% 
  group_by(name) %>% 
  summarise(mindiff = min(diff)) %>%
  arrange(desc(mindiff)) %>%
  kable()
```

## GWS has played two draws in a row. How many times has this happened in the same season?
```{r q5}
games %>% 
  mutate(Date = as_date(Date, "%d-%b-%Y")) %>% 
  select(team1, team2, round, Date) %>% 
  gather(team, name, -round, -Date) %>% 
  arrange(name, Date) %>% 
  mutate(diff = Date - lag(Date)) %>% 
  filter(diff >= 0, diff <= 8) %>% 
  arrange(name, diff) %>%
  kable()
```

## Which season has had the greatest number of drawn games?
```{r q6}
games %>% 
  count(Year) %>% 
  arrange(desc(n)) %>%
  slice(1) %>%
  kable()
```

## What was the highest score in a drawn game?
```{r q7}
games %>% 
  mutate(total = as.numeric(sapply(strsplit(score1, "\\."), function(x) x[3]))) %>%
  arrange(desc(total)) %>% 
  slice(1) %>% kable()
```

## How many drawn finals have there been?
```{r q8}
games %>%
  filter(grepl("F", round)) %>%
  kable()
```

## Which two clubs were involved in the first draw in the 1897 season?
```{r q9}
games %>% 
  slice(1) %>%
  kable()
```

## In what season did South Melbourne make the finals because they had drawn a game?
 [1977](http://australianfootball.com/articles/view/Drawing%2Bthe%2Bbest%2Bconclusion/177)
