---
title: "Returning AFL coaches"
author: "Neil Saunders"
date: "`r Sys.time()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)
library(dplyr)
library(stringr)
library(tweetrmd)
library(rvest)
library(knitr)
library(kableExtra)
```

# Introduction

We might ask the question:

```{r fetch-tweet, echo=FALSE}
include_tweet("https://twitter.com/crow_data_sci/status/1580744416561856512")
```

Let's try to find out using data from [AFL Tables](https://afltables.com/afl/afl_index.html).

# Data processing

# Coaching data

Coach names and links to their individual data are located [in this HTML table](https://afltables.com/afl/stats/coaches/coaches_idx.html).

We can fetch and process it like so:

```{r get-coaches}
p <- read_html("https://afltables.com/afl/stats/coaches/coaches_idx.html")
coaches <- tibble(name = p %>% 
                    html_nodes("table a") %>% 
                    html_text(), 
                  link = p %>% 
                    html_nodes("table a") %>% 
                    html_attr("href")) %>% 
  slice(1:378)
```

```{r preview-coaches}
coaches %>% 
  head(10) %>% 
  kbl(format = "simple") %>% 
    kable_styling(bootstrap_options = c("striped"))
```

## Counting clubs

The function `count_coaches_teams()` does the following:

1. Fetches the data page for a coach
1. Extracts the names of each club coached and the years that the coach was at the club
1. Determines whether the coach (1) left to coach another club and then returned or (2) had a break of one or more seasons then returned to the same club
1. Returns the results

```{r count-coaches-teams}
count_coaches_teams <- function(coachhtml) {
  coachurl <- paste0("https://afltables.com/afl/stats/coaches/", coachhtml) %>% 
    URLencode()
  coach_data <- coachurl %>% 
    read_html() %>% 
    html_table(header = FALSE) %>%
    .[[1]] %>%
    filter(str_detect(X1, "\\d{4}")) %>%
    select(1:2) %>% 
    setNames(c("year", "team"))
  
  team_count <- coach_data %>% 
    mutate(from = lag(team, default = "None"), 
           to = team,
           year = as.numeric(year),
           flag_same_club = ifelse(from == to, 1, 0),
           flag_gap_year = ifelse(year - lag(year) > 1, 1, 0))
  
  diff_team_count <- team_count %>% 
    filter(flag_same_club == 0) %>% 
    count(to) %>% 
    filter(n > 1) %>% 
    mutate(link = coachhtml,
           type = "return from other club")
  
  same_team_count <- team_count %>% 
    filter(flag_same_club == 1,
           flag_gap_year == 1) %>% 
    count(to) %>% 
    mutate(link = coachhtml,
           type = "return to same club after break")
  
  output <- bind_rows(diff_team_count, same_team_count)
  output
}
```

## Worked examples

### Mick Malthouse

Coached 4 different clubs, consecutively with no breaks. Therefore `count_coaches_teams` returns no rows.

```{r ex01}
  coach_data <- "https://afltables.com/afl/stats/coaches/Mick_Malthouse.html" %>% 
    read_html() %>% 
    html_table(header = FALSE) %>%
    .[[1]] %>%
    filter(str_detect(X1, "\\d{4}")) %>%
    select(1:2) %>% 
    setNames(c("year", "team"))

coach_data

  team_count <- coach_data %>% 
    mutate(from = lag(team, default = "None"), 
           to = team,
           year = as.numeric(year),
           flag_same_club = ifelse(from == to, 1, 0),
           flag_gap_year = ifelse(year - lag(year) > 1, 1, 0))

team_count

mm <- count_coaches_teams("Mick_Malthouse.html")

mm
```

### Jock McHale

Coached 1 club continuously with no breaks. Therefore `count_coaches_teams` returns no rows.

```{r ex02}
  coach_data <- "https://afltables.com/afl/stats/coaches/Jock_McHale.html" %>% 
    read_html() %>% 
    html_table(header = FALSE) %>%
    .[[1]] %>%
    filter(str_detect(X1, "\\d{4}")) %>%
    select(1:2) %>% 
    setNames(c("year", "team"))

coach_data

  team_count <- coach_data %>% 
    mutate(from = lag(team, default = "None"), 
           to = team,
           year = as.numeric(year),
           flag_same_club = ifelse(from == to, 1, 0),
           flag_gap_year = ifelse(year - lag(year) > 1, 1, 0))

team_count

jm <- count_coaches_teams("Jock_McHale.html")

jm
```

### Allan Jeans

Coached 3 clubs, the first continuously, the second with a break.

```{r ex03}
  coach_data <- "https://afltables.com/afl/stats/coaches/Allan_Jeans.html" %>% 
    read_html() %>% 
    html_table(header = FALSE) %>%
    .[[1]] %>%
    filter(str_detect(X1, "\\d{4}")) %>%
    select(1:2) %>% 
    setNames(c("year", "team"))

coach_data

  team_count <- coach_data %>% 
    mutate(from = lag(team, default = "None"), 
           to = team,
           year = as.numeric(year),
           flag_same_club = ifelse(from == to, 1, 0),
           flag_gap_year = ifelse(year - lag(year) > 1, 1, 0))

team_count

aj <- count_coaches_teams("Allan_Jeans.html")

aj
```

### David Parkin

Coached 3 clubs, returning to Carlton after a spell at Fitzroy.

```{r ex04}
  coach_data <- "https://afltables.com/afl/stats/coaches/David_Parkin.html" %>% 
    read_html() %>% 
    html_table(header = FALSE) %>%
    .[[1]] %>%
    filter(str_detect(X1, "\\d{4}")) %>%
    select(1:2) %>% 
    setNames(c("year", "team"))

coach_data

  team_count <- coach_data %>% 
    mutate(from = lag(team, default = "None"), 
           to = team,
           year = as.numeric(year),
           flag_same_club = ifelse(from == to, 1, 0),
           flag_gap_year = ifelse(year - lag(year) > 1, 1, 0))

team_count

dp <- count_coaches_teams("David_Parkin.html")

dp
```

These examples capture the possibilities.

# Putting it all together

Let's apply the function to all coaches.

```{r process-all-coaches}
teams_list <- lapply(coaches$link, count_coaches_teams)

teams_list_df <- teams_list %>% 
  bind_rows() %>% 
  mutate(coach = gsub("(_|\\.html)", " ", link) %>% 
           trimws()) %>% 
  select(coach, returning_to = to, count = n, return_type = type)
```

The full list of coaches who returned to a club, either after a break or after time coaching another club.

There are `r nrow(teams_list_df)` instances, involving `r n_distinct(teams_list_df$coach)` coaches, returning to `r n_distinct(teams_list_df$returning_to)` clubs.

```{r finaloutput}
teams_list_df %>% 
  arrange(coach) %>% 
  kbl(format = "simple") %>% 
    kable_styling(bootstrap_options = c("striped"))
```

