---
title: "Sydney Swans Home Game Crowds"
author: "Neil Saunders"
date: "19/05/2015"
output: 
  html_document: 
    keep_md: yes
---

## Introduction
This document compares crowds for home games at the Sydney Cricket Ground ("S.C.G.") versus ANZ Stadium (formerly known as Stadium Australia, "Stad. Aust.").

## 1. Getting the data
Data for all Swans games by season are obtained from [AFL Tables](http://afltables.com/afl/teams/swans/allgames.html). The functions used are shown at the end of this document.

```{r warning=FALSE, message=FALSE}
setwd("~/Dropbox/projects/github_projects/aflstats/code/R")
source("scripts/afltables.R")

swans <- allGames("swans")
swans$DateTime <- allGamesDateTime(swans)
swans$date     <- allGamesDate(swans)
swans$Year     <- allGamesYear(swans)

# get the home games since moving to Sydnet
swans.h <- subset(swans, T == "H" & Year > 1981)
table(swans.h$Venue)
```

## 2. Plotting the data
We start with a basic scatterplot coloured by Venue.

```{r tidy=TRUE, fig.height=7, fig.width=9}
library(ggplot2)
ggplot(swans.h) + geom_point(aes(date, Crowd, color = Venue)) + theme_bw() + scale_color_manual(values = c("darkorange", "cornflowerblue"))+ labs(x = "Date", title = "Sydney Swans Home Game Crowds")
```

Clearly there are two "eras" of crowd numbers: small crowds pre-1996 (with spikes in 1986-1987) and larger crowds post-1996. We can indicate the median crowds for those two time periods, for the SCG and ANZ Stadium.

```{r tidy=TRUE, fig.height=7, fig.width=9}
ggplot(swans.h) + geom_point(aes(date, Crowd, color = Venue)) + theme_bw() + scale_color_manual(values = c("darkorange", "cornflowerblue")) + geom_segment(aes(x = as.Date("1982-01-01", "%Y-%m-%d"), xend = as.Date("1995-12-31", "%Y-%m-%d"), y = median(subset(swans.h, Venue == "S.C.G." & Year < 1996)$Crowd), yend = median(subset(swans.h, Venue == "S.C.G." & Year < 1996)$Crowd)), color = "darkorange", linetype = "dashed") + geom_segment(aes(x = as.Date("1996-01-01", "%Y-%m-%d"), xend = as.Date("2015-12-31", "%Y-%m-%d"), y = median(subset(swans.h, Venue == "S.C.G." & Year >= 1996)$Crowd), yend = median(subset(swans.h, Venue == "S.C.G." & Year >= 1996)$Crowd)), color = "darkorange") + geom_segment(aes(x = as.Date("2002-01-01", "%Y-%m-%d"), xend = as.Date("2015-12-31", "%Y-%m-%d"), y = median(subset(swans.h, Venue == "Stad. Aust." & Year >= 2002)$Crowd), yend = median(subset(swans.h, Venue == "Stad. Aust." & Year >= 2002)$Crowd)), color = "cornflowerblue") + labs(x = "Date", title = "Sydney Swans Home Game Crowds\nshowing medians for SCG (1982-1995 and 1996-2015) and ANZ (2002-2015")
```

Finally, we can try to superimpose a trend for crowds at ANZ Stadium.

```{r tidy=TRUE, fig.height=7, fig.width=9}
ggplot(swans.h) + geom_point(aes(date, Crowd, color = Venue)) + theme_bw() + scale_color_manual(values = c("darkorange", "cornflowerblue")) + geom_segment(aes(x = as.Date("1982-01-01", "%Y-%m-%d"), xend = as.Date("1995-12-31", "%Y-%m-%d"), y = median(subset(swans.h, Venue == "S.C.G." & Year < 1996)$Crowd), yend = median(subset(swans.h, Venue == "S.C.G." & Year < 1996)$Crowd)), color = "darkorange", linetype = "dashed") + geom_segment(aes(x = as.Date("1996-01-01", "%Y-%m-%d"), xend = as.Date("2015-12-31", "%Y-%m-%d"), y = median(subset(swans.h, Venue == "S.C.G." & Year >= 1996)$Crowd), yend = median(subset(swans.h, Venue == "S.C.G." & Year >= 1996)$Crowd)), color = "darkorange") + geom_smooth(data = subset(swans.h, Venue == "Stad. Aust."), aes(x = date, y = Crowd)) + labs(x = "Date", title = "Sydney Swans Home Game Crowds\nshowing SCG medians (1982-1995 and 1996-2015) and ANZ trend")
```

## 3. Summary
In general, home game crowds have been larger at ANZ Stadium compared with the SCG. We can speculate that this is due in large part to marketing; games at ANZ Stadium are promoted as "big games" (finals, not included in this analysis or games featuring high-quality opposition).

However, in recent years crowds have declined to levels comparable with the SCG. This may be part of the reason behind [the decision](www.afl.com.au/news/2014-11-24/swans-set-to-quit-homebush) to host all home games at the SCG after 2016.

### Functions used

```{r eval=FALSE, tidy=TRUE}
# all games given a team name
allGames <- function(team) {
  require(XML)
  require(plyr)
  games <- readHTMLTable(paste("http://afltables.com/afl/teams/", team, "/allgames.html", sep = ""), stringsAsFactors = FALSE)
  # fix for 2 "Scoring" columns
  for(i in 1:length(games)) {
    colnames(games[[i]])[c(4,6)] <- c("Scoring.F", "Scoring.A")
    }
  games <- ldply(games, rbind)
  games$.id <- NULL
  games$F <- as.numeric(games$F)
  games$A <- as.numeric(games$A)
  games$Crowd <- as.numeric(games$Crowd)
  return(games)
}

# convert game date to POSIXlt/POSIXt class
allGamesDateTime <- function(games) {
  datetime <- strptime(games$Date, format = "%d-%b-%Y %I:%M %p")
  return(datetime)
}

# convert game datetime to Date class
allGamesDate <- function(games) {
  datetime <- allGamesDateTime(games)
  date     <- as.Date(datetime)
  return(date)
}

# convert game date to year
allGamesYear <- function(games) {
  datetime <- allGamesDateTime(games)
  year     <- datetime$year + 1900
  return(year)
}
```