---
title: "Sydney Swans Home Game Crowds"
author: "Neil Saunders"
date: "19/05/2015"
output:
  github_document:
    toc: true
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)
library(tidyverse)
library(rvest)
library(lubridate)

theme_set(theme_bw())

swans <- read_html("https://afltables.com/afl/teams/swans/allgames.html") %>% 
  html_table(fill = TRUE, header = FALSE) %>% 
  bind_rows() %>% 
  filter(!X1 %in% c("Rnd", "Totals", "Averages")) %>% 
  select(Type = X2, Venue = X11, Crowd = X12, Date = X13) %>% 
  mutate(Crowd = as.numeric(Crowd), 
         Date = as.Date(Date, "%a %d-%b-%Y %I:%M %p"),
         Year = year(Date)) %>% 
  filter(Type == "H", Date > as.Date("1981-12-31"))
```


# Introduction
This document compares crowds for home games at the Sydney Cricket Ground ("S.C.G.") versus ANZ Stadium (formerly known as Stadium Australia, "Stad. Aust.").

# Getting the data
Data for all Swans games by season are obtained from [AFL Tables](https://afltables.com/afl/teams/swans/allgames.html).

# Plotting the data
We start with a basic scatterplot coloured by Venue.

```{r fig.height=7, fig.width=9}
swans %>%
  ggplot() + 
    geom_point(aes(Date, Crowd, color = Venue)) + 
    scale_color_manual(values = c("darkorange", "cornflowerblue")) + 
    labs(x = "Date", title = "Sydney Swans Home Game Crowds")
```

Clearly there are two "eras" of crowd numbers: small crowds pre-1996 (with spikes in 1986-1987) and larger crowds post-1996. We can indicate the median crowds for those two time periods, for the SCG and ANZ Stadium.

```{r fig.height=7, fig.width=9}
swans %>%
  ggplot() + 
    geom_point(aes(Date, Crowd, color = Venue)) + 
    scale_color_manual(values = c("darkorange", "cornflowerblue")) + 
    geom_segment(aes(x = as.Date("1982-01-01", "%Y-%m-%d"), 
                     xend = as.Date("1995-12-31", "%Y-%m-%d"), 
                     y = median(subset(swans, Venue == "S.C.G." & Year < 1996)$Crowd), 
                     yend = median(subset(swans, Venue == "S.C.G." & Year < 1996)$Crowd)), 
                 color = "darkorange", linetype = "dashed") + 
    geom_segment(aes(x = as.Date("1996-01-01", "%Y-%m-%d"), 
                     xend = as.Date("2015-12-31", "%Y-%m-%d"), 
                     y = median(subset(swans, Venue == "S.C.G." & Year >= 1996)$Crowd), 
                     yend = median(subset(swans, Venue == "S.C.G." & Year >= 1996)$Crowd)), 
                color = "darkorange") + 
    geom_segment(aes(x = as.Date("2002-01-01", "%Y-%m-%d"), 
                     xend = as.Date("2015-12-31", "%Y-%m-%d"), 
                     y = median(subset(swans, Venue == "Stad. Aust." & Year >= 2002)$Crowd), 
                     yend = median(subset(swans, Venue == "Stad. Aust." & Year >= 2002)$Crowd)), 
                 color = "cornflowerblue") + 
    labs(x = "Date", 
         title = "Sydney Swans Home Game Crowds",
         subtitle = "showing medians for SCG (1982-1995 and 1996-2015) and ANZ (2002-2015")
```

Finally, we can try to superimpose a trend for crowds at ANZ Stadium.

```{r fig.height=7, fig.width=9}
swans %>%
  ggplot() + 
    geom_point(aes(Date, Crowd, color = Venue)) + 
    scale_color_manual(values = c("darkorange", "cornflowerblue")) + 
    geom_segment(aes(x = as.Date("1982-01-01", "%Y-%m-%d"), 
                     xend = as.Date("1995-12-31", "%Y-%m-%d"),
                     y = median(subset(swans, Venue == "S.C.G." & Year < 1996)$Crowd), 
                     yend = median(subset(swans, Venue == "S.C.G." & Year < 1996)$Crowd)), 
                 color = "darkorange", linetype = "dashed") + 
    geom_segment(aes(x = as.Date("1996-01-01", "%Y-%m-%d"), 
                     xend = as.Date("2015-12-31", "%Y-%m-%d"), 
                     y = median(subset(swans, Venue == "S.C.G." & Year >= 1996)$Crowd), 
                     yend = median(subset(swans, Venue == "S.C.G." & Year >= 1996)$Crowd)), 
                 color = "darkorange") + 
    geom_smooth(data = subset(swans, Venue == "Stad. Aust."), aes(x = Date, y = Crowd)) + 
    labs(x = "Date", 
         title = "Sydney Swans Home Game Crowds", 
         subtitle = "showing SCG medians (1982-1995 and 1996-2015) and ANZ trend")
```

# Summary
In general, home game crowds have been larger at ANZ Stadium compared with the SCG. We can speculate that this is due in large part to marketing; games at ANZ Stadium are promoted as "big games" (finals, not included in this analysis or games featuring high-quality opposition).

However, in recent years crowds have declined to levels comparable with the SCG. This may be part of the reason behind [the decision](www.afl.com.au/news/2014-11-24/swans-set-to-quit-homebush) to host all home games at the SCG after 2016.
