---
title: "Sydney Swans/South Melbourne greatest (and worst) Q4 turnarounds"
author: "Neil Saunders"
date: "compiled `r Sys.time()`"
output:
  github_document:
    toc: true
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)
library(tidyverse)
library(rvest)
library(lubridate)
library(pander)

theme_set(theme_bw())

swans <- read_html("https://afltables.com/afl/teams/swans/allgames.html") %>% 
  html_table(fill = TRUE, header = FALSE) %>% 
  bind_rows() %>% 
  filter(!X1 %in% c("Rnd", "Totals", "Averages")) %>%
  rename(rnd = X1, type = X2, opponent = X3, scoring_for = X4, points_for = X5, 
         scoring_against = X6, points_against = X7, result = X8, margin = X9, wdl = X10, 
         venue = X11, crowd = X12, date = X13) %>% 
  mutate(points_for = as.numeric(points_for), 
         points_against = as.numeric(points_against), 
         margin = as.numeric(margin), 
         crowd = as.numeric(crowd), 
         date = dmy_hm(date))
```

# Introduction

How many times have the Swans come from behind going into the last quarter to win? Or been in front, but then gone on to lose? In each case, what were the biggest swings in the scoreline?

# Getting the data

We download and process Swans game data from [AFL Tables](http://afltables.com).

There are currently `r nrow(swans)` of them.

# Analysis
## Come from behind at Q3 to win
We select the subset of cases where the margin is negative at the end of Q3, but the result is a win. Then we sort by the difference between Q3 and Q4 margin.

The top 10:
```{r}
swans_win <- swans %>% 
  mutate(q3gfor = str_match(scoring_for, "^(\\d+\\.\\d+\\s+){2}(\\d+)")[, 3], 
         q3bfor = str_match(scoring_for, "^(\\d+\\.\\d+\\s+){2}\\d+\\.(\\d+)")[, 3], 
         q3gagainst = str_match(scoring_against, "^(\\d+\\.\\d+\\s+){2}(\\d+)")[, 3], 
         q3bagainst = str_match(scoring_against, "^(\\d+\\.\\d+\\s+){2}\\d+\\.(\\d+)")[, 3], 
         q3margin = (6 * as.numeric(q3gfor) + as.numeric(q3bfor)) - (6 * as.numeric(q3gagainst) + as.numeric(q3bagainst)), 
         turnaround = margin - q3margin) %>% 
  filter(result == "W", 
         q3margin < 0)

swans_win %>% 
  top_n(10, turnaround) %>% 
  arrange(desc(turnaround)) %>%
  select(date, rnd, type, opponent, q3margin, margin) %>%
  pander()
```

They've come from behind in Q3 to win `r nrow(swans_win)` times.

## Slip from in front at Q3 to lose
The procedure is very similar, except that we select the cases where margin is positive at the end of Q3, but the result is a loss.

```{r}
swans_loss <- swans %>% 
  mutate(q3gfor = str_match(scoring_for, "^(\\d+\\.\\d+\\s+){2}(\\d+)")[, 3], 
         q3bfor = str_match(scoring_for, "^(\\d+\\.\\d+\\s+){2}\\d+\\.(\\d+)")[, 3], 
         q3gagainst = str_match(scoring_against, "^(\\d+\\.\\d+\\s+){2}(\\d+)")[, 3], 
         q3bagainst = str_match(scoring_against, "^(\\d+\\.\\d+\\s+){2}\\d+\\.(\\d+)")[, 3], 
         q3margin = (6 * as.numeric(q3gfor) + as.numeric(q3bfor)) - (6 * as.numeric(q3gagainst) + as.numeric(q3bagainst)), 
         turnaround = margin - q3margin) %>% 
  filter(result == "L", q3margin > 0)

swans_loss %>% 
  top_n(10, abs(turnaround)) %>% 
  arrange(turnaround) %>%
  select(date, rnd, type, opponent, q3margin, margin) %>% 
  pander()
```

They've thrown it away in the last quarter `r nrow(swans_loss)` times.
